---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# envvar

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/envvar)](https://CRAN.R-project.org/package=envvar)
[![R-CMD-check](https://github.com/briandconnelly/envvar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/briandconnelly/envvar/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/briandconnelly/envvar/branch/main/graph/badge.svg)](https://app.codecov.io/gh/briandconnelly/envvar?branch=main)
<!-- badges: end -->

Environment variables are a powerful tool that enable your code to react to its environment.
However, two common design choices are a frequent source of friction.
First, unlike most other "getter"-type functions, functions that retrieve values from environment variable typically fail silently.
Because of this, additional code to check whether the environment variable was set are needed.
Second, values are almost always returned as strings.
This is understandable, but programmers often use environment variables to store a wide variety of data types from numbers to timestamps to URLs.
In this case, additional code is required to coerce those strings into their intended format.
For frequent users of environment variables, writing all this extra code is unpleasant and time consuming.

envvar takes a slightly opinionated perspective to make working with environment variables easier and more consistent.
Unless a default value is explicitly given, `envvar_get()` raises an error.
For example, let's say our code depends on environment variable called `NUM_CPUS`.
In base R, we have to first get the value using `Sys.getenv()` and then see whether the result is the empty string (not `NA` like you might expect):

```{r base notset,error=TRUE}
num_cpus <- Sys.getenv("NUM_CPUS")

if (identical(num_cpus, "")) {
  stop("I need `NUM_CPUS` to be set!")
}
```

envvar's `envvar_get()` will just fail if `NUM_CPUS` isn't set:

```{r envvar notset,error=TRUE}
library(envvar)

envvar_get("NUM_CPUS")
```

If a reasonable default is known, it can be supplied via the `default` argument.
envvar prints a message, though, so you know that it's using a default rather than a value specified in the environment.

```{r envvar notset default}
envvar_get("NUM_CPUS", default = 12)
```


## Speaking Native Types

```{r include=FALSE}
Sys.setenv("NUM_CPUS" = 8)
```

Let's say our `NUM_CPUS` environment variable is set.
For this example, it's set to 8.
Because `Sys.getenv()` returns strings, we can't immediately treat it like the integer that it is.

```{r base nocoerce,error=TRUE}
Sys.getenv("NUM_CPUS") / 2

num_cpus / 2
```

envvar includes several helper functions that return commonly-used data types as their proper type.
Here, we'll use `envvar_get_integer()` to get `NUM_CPUS` and return it as an integer.

```{r envvar nocoerce,error=TRUE}
envvar::envvar_get_integer("NUM_CPUS") / 2
```

envvar can handle numbers, logical values, version numbers, URLs, timestamps, UUIDs, IP addresses, and more.
We'll work with dates in the next example.


## Validation

envvar's `envvar_get` functions can also apply validation logic.
For this example, let's set an environment variable called `LAUNCH_DATE` that absolutely, positively must be in the future.
For now, let's set it to a date in the past.

```{r set launch date past}
envvar_set("LAUNCH_DATE" = "1969-07-16")
```

To read `LAUNCH_DATE` and ensure that it is in the future, we can supply a function to `envvar_get_date()` that checks the value.
If this function returns `FALSE`, an error is raised.

```{r get launch date,error=TRUE}
envvar_get_date("LAUNCH_DATE", validate = \(x) x > Sys.Date())
```

Let's try that again:
```{r get good launch date,error=TRUE}
envvar_set("LAUNCH_DATE" = "2028-08-28")

envvar_get_date("LAUNCH_DATE", validate = \(x) x > Sys.Date())
```


## Installation

envvar is still changing quickly, so it's not quite ready for CRAN.
If you’d like to try out the development version, you can install directly from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("briandconnelly/envvar")
```


## Related Packages

- [dotenv](https://github.com/gaborcsardi/dotenv) package for loading environment variables from `.env` files
- [config](https://rstudio.github.io/config/) package for defining and using multiple environments
- [options](https://dgkf.github.io/options/) package for defining and using R package options, another way of adding flexibility to your code

