---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# envvar

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/envvar)](https://CRAN.R-project.org/package=envvar)
[![R-CMD-check](https://github.com/briandconnelly/envvar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/briandconnelly/envvar/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/briandconnelly/envvar/branch/main/graph/badge.svg)](https://app.codecov.io/gh/briandconnelly/envvar?branch=main)
<!-- badges: end -->

Environment variables are a common way to alter how your code behaves depending on context.
They're represented by name-value pairs, and the value is always a string.
Because of this, most code that uses environment variables must validate those values to make sure they're reasonable and then transform them into a more usable type.

envvar makes working with environment variables easier and more consistent.
It enables you to transform and validate values, making them immediately usable in your code.
Aside from allowing you to define your own transformation and validation functions, envvar comes with its own set of helper functions that handle common data types such as integers, lists, URLs, IP addresses, and more.

For example, pretend we're using an environment variable called `IMPORTANT_SERVER` to store the URL of an important server.
We can use envvar's `envvar_get_url()` function to receive its value and convert it into an [httr2] `url` object:

```{r set_url}
library(envvar)

envvar_set("IMPORTANT_SERVER" = "https://192.168.1.12")
```

```{r get_url}
library(envvar)

prod_server_url <- envvar_get_url("IMPORTANT_SERVER")
prod_server_url
```

Just to be safe, we can set a default value in case someone forgot to set the value:

```{r use_default}
envvar_unset("IMPORTANT_SERVER")

prod_server_url <- envvar_get_url("IMPORTANT_SERVER", default = "https://my_important_server.biz")
prod_server_url
```

We can also add a validation check to ensure it's using https:

```{r validate_url}
envvar_set("IMPORTANT_SERVER" = "https://192.168.1.12")

prod_server_url <- envvar_get_url(
  "IMPORTANT_SERVER",
  default = "https://my_important_server.biz",
  validate = \(x) x$scheme == "https"
)
prod_server_url
```



## Installation

If you’d like to try out the development version of envvar, you can install directly from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("briandconnelly/envvar")
```
